### 2020-04-28 manuscript/manuscript.rmd
fixed a few formatting issues to reknit for pdf and html. Set html to default output and removed other output formats.

---
bibliography: manuscript.bib
csl: nature.csl
output:
  pdf_document: default
  word_document: default
  html_document: default
  github_document: default
---

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$ , the rate of evolution, and the amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the rate at which one amino acid changes to another (*e.g.* Poisson/F81, WAG or LG, and GTR). The rate of exchange can be further scaled, for example with a Gamma parameter that accommodates rate variation across sites. Heterogeneous equilibrium approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components– CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). Moreover, the Gamma-rate heterogeneity is used in almost every analysis conducted here. If there are no modifiers, then it is implied that site homogeneous models are used. A detailed description of each component is provided in Supplementary Results.

@@

We first used the Metazoa BUSCO genes [@waterhouse2017busco] to evaluate gene composition from all data matrices and the result is unexpected (Extended Data Fig. 2A). The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of assemblies from genomic data [@waterhouse2017busco]. The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices. However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Extended Data Fig. 2B). Moreover, we identified number of ribosomal RNA genes in each data matrix (Extended Data Fig. 2C) and found that Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig. 2D). We also annotated each gene partition against SwissProt [@boeckmann2003swiss] and Gene Ontology databases, and results are shown in Supplementary Table 1.
Moreover, our network analyses show that surprisingly little overlap of gene samplings across all the data matrices from different studies (Fig. 2). Thus, these results strongly suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.

@@

In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. 5B; Supplementary Table 4). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Figs. 5A, C). We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4). The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.

@@

We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories in the representative data matrices (Fig. XX). Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar evolutioanry models. Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories (Fig. XX). More importantly, we are able to identify the transition of topology between the two alternative hypotheses in these three matrices. For example, the major transition occurred at 60 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the Whelan2017_strict matrix. Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than the site-homogeneous models [@whelan2016let].

@@

For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson-CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above. For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority‐rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks. We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.

@@

![](figures/Figure_Pairwise_heatmap_animal_root.png)
**Extended Data Fig. XXpairwise_heatmap.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance. (A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses. (B) Pairwise comparisons across genes that supported Porifera-sister in different analyses. The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation. The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].

@@

# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
#                    widths = unit(c(4, 4), "cm"),
#                    heights = unit(c(2, 5), c("in", "lines")))
#
# grid.newpage()
# grid.draw(fig5)
```
**Extended Data Fig. XXphylogetic signal** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice. (A). Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Philippe2009 matrix. (B). Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Ryan2013_est matrix. (C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix. The (\Delta$GLS\) values for the genes across each data matrix are provided in Supplementary Table 4.

@@

**Extended Data Fig. XXrandom_recoding.** (A) The two alternative hypotheses for deep animal relationships considered here. Relationships that are not part of these hypotheses are shown as unresolved polytomies. (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses. These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03). Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point). For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding. This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.

@@

> The resulting phylogeny yields two significant conclusions reviving old views that have been challenged in the molecular era: (1) that the sponges (Porifera) are monophyletic and not paraphyletic as repeatedly proposed, thus undermining the idea that ancestral metazoans had a sponge-like body plan; (2) that the most likely position for the ctenophores is together with the cnidarians in a ‘‘coelenterate’’ clade.

>>>>

---
bibliography: manuscript.bib
csl: nature.csl
output:
  html_document: default
  pdf_document: default
  word_document: default
  github_document: default
---

@@

Models of molecular evolution have several components that each consider different aspects of the evolutionary process. The models that have been used to model protein evolution in studies of the deep phylogeny have largely differed according to three components: the exchangeability matrix $E$ , the rate of evolution, and the amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the rate at which one amino acid changes to another (*e.g.* Poisson/F81, WAG or LG, and GTR). The rate of exchange can be further scaled, for example with a Gamma parameter that accommodates rate variation across sites. Heterogeneous equilibrium approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). Moreover, the Gamma-rate heterogeneity is used in almost every analysis conducted here. If there are no modifiers, then it is implied that site homogeneous models are used. A detailed description of each component is provided in Supplementary Results.

@@

We first used the Metazoa BUSCO genes [@waterhouse2017busco] to evaluate gene composition from all data matrices and the result is unexpected (Extended Data Fig. 2A). The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of assemblies from genomic data [@waterhouse2017busco]. The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices. However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Extended Data Fig. 2B). Moreover, we identified number of ribosomal RNA genes in each data matrix (Extended Data Fig. 2C) and found that Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig. 2D). We also annotated each gene partition against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases, and results are shown in Supplementary Table 1.
Moreover, our network analyses show that surprisingly little overlap of gene samplings across all the data matrices from different studies (Fig. 2). Thus, these results strongly suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.

@@

In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal ($\Delta$lnL<2) (Fig. 5B; Supplementary Table 4). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Figs. 5A, C). We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4). The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.

@@

We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories in the representative data matrices (Fig. XX). Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar evolutioanry models. Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories (Fig. XX). More importantly, we are able to identify the transition of topology between the two alternative hypotheses in these three matrices. For example, the major transition occurred at 60 -- 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the Whelan2017_strict matrix. Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than the site-homogeneous models [@whelan2016let].

@@

For several Choanozoa matrices indicated strong support for the hypothesis that sponges are the sister group to the remaining Metazoa using the Poisson-CAT model, bayesian inference with Poisson-CAT model was also conducted to Holozoa and Opisthokonta data matrices with the same settings as above. For all the analyses with Poisson+CAT models in PhyloBayes, two independent chains were sampled every generation. Tracer plots of MCMC runs were visually inspected in Tracer v1.6 to assess stationarity and appropriate burn-in. Chains were considered to have reached convergence when the maxdiff statistic among chains was below 0.3 (as measured by bpcomp) and effective sample size > 50 for each parameter (as measured by tracecomp). A 50% majority-rule consensus tree was computed with bpcomp, and nodal support was estimated by posterior probability. For those matrices that were not converged, PhyloBayes analyses were run for at least two weeks. We also summarized the average number of substitutional categories inferred for each PhyloBayes analysis using Tracer.

@@

![](figures/Figure_Pairwise_heatmap_animal_root.png)
**Extended Data Fig. XXpairwise_heatmap.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance. (A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses. (B) Pairwise comparisons across genes that supported Porifera-sister in different analyses. The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and -1 is negative linear correlation. The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].

@@

# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
#                    widths = unit(c(4, 4), "cm"),
#                    heights = unit(c(2, 5), c("in", "lines")))
#
# grid.newpage()
# grid.draw(fig5)
```
**Extended Data Fig. XXphylogetic signal** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice. (A). Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Philippe2009 matrix. (B). Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Ryan2013_est matrix. (C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix. The $\Delta$GLS values for the genes across each data matrix are provided in Supplementary Table 4.

@@

**Extended Data Fig. XXrandom_recoding.** (A) The two alternative hypotheses for deep animal relationships considered here. Relationships that are not part of these hypotheses are shown as unresolved polytomies. (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses. These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 - random-03). Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point). For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding. This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.

@@

> The resulting phylogeny yields two significant conclusions reviving old views that have been challenged in the molecular era: (1) that the sponges (Porifera) are monophyletic and not paraphyletic as repeatedly proposed, thus undermining the idea that ancestral metazoans had a sponge-like body plan; (2) that the most likely position for the ctenophores is together with the cnidarians in a 'coelenterate' clade.
